cardinalR: Generating interesting high-dimensional data structures

### 2025-09-20 paper-cardinalR.Rmd
Updated branching section

### Branching

@@

#### `gen_expbranches()`

The `gen_expbranches(n, p, k)` function generates a dataset of $n$ points forming $k$ exponential branches in $2\text{-}D$, with optional noise dimensions to embed the structure in $p\text{-}D$ (`r knitr::asis_output(if (knitr::is_html_output()) { "Figure \\@ref(fig:branch) a"} else { "Figure \\@ref(fig:branch-proj) a"})`). These branches grow in opposite directions, producing a radiating curvilinear structure from a central region.

Each branch $i$ is constructed using $X_1 \sim U(-2, 2)$ and $X_2 = \exp(\pm s_iX_1) + \epsilon$, where: $\epsilon \sim U(0, 0.1)$ adds local jitter, $s_i \in [0.5, 2]$ is randomly sampled for each branch. The sign of the exponent alternates between branches: odd-numbered branches have a negative exponent (decaying pattern), and even-numbered branches have a positive exponent (growing pattern), producing mirror-symmetric curves with varying steepness. For $p > 2$, Gaussian noise $X_j \sim N(0, 0.1^2)$ is added to embed the $2\text{-}D$ branches into $p\text{-}D$, where $j = 3, \dots, p$.

@@

#### `gen_linearbranches()`

The `gen_linearbranches(n, p, k)` function generates a dataset of $n$ points forming $k$ approximately linear branches in $p\text{-}D$ (`r knitr::asis_output(if (knitr::is_html_output()) { "Figure \\@ref(fig:branch) b"} else { "Figure \\@ref(fig:branch-proj) b"})`). The core structure lies in the first two dimensions and additional dimensions carry Gaussian noise.

The `gen_linearbranches(n, p, k)` function generates a dataset of $n$ points in $p\text{-}D$ forming $k$ linear branches. Each branch is a line segment with added jitter to simulate noise. Branches $1$ and $2$ are initialized with fixed slopes and intercepts: branch $1$ has slope $0.5$ and intercept $0$, with $X_1 \sim U(-2, 8)$ and $X_2 = 0.5 X_1 + \epsilon$, $\epsilon \sim U(0, 0.5)$; branch $2$ has slope $-0.5$ and intercept $0$, with $X_1 \sim U(-6, 2)$ and $X_2 = -0.5 X_1 + \epsilon$, $\epsilon \sim U(0, 0.5)$.

Branches $3$ to $k$ are added iteratively. Each new branch starts at a location outside predefined exclusion zones to avoid overlap with the first two branches. Specifically, $X_1$ starting points are sampled outside the intervals $[-8, -7]$, $[-2, 2]$, and $[7, 8]$, and $X_2$ starting points are sampled outside $[7, 8]$. For each new branch $i$, $X_1$ values are drawn over a short interval $[x_{\text{start}}, x_{\text{start}} + 1]$, and $X_2$ values are calculated as $X_2 = s_i (X_1 - x_{\text{start}}) + y_{\text{start}} + \epsilon$, where $s_i$ is a randomly chosen slope from a filtered sequence and $\epsilon \sim U(0, 0.2)$. For $p > 2$, Gaussian noise $X_j \sim N(0, 0.05^2)$ is added to embed the $2\text{-}D$ branches into $p\text{-}D$, for $j = 3, \dots, p$.

#### `gen_curvybranches()`

The `gen_curvybranches(n, p, k)` function generates a dataset of $n$ points forming $k$ curvilinear branches embedded in $p\text{-}D$ (`r knitr::asis_output(if (knitr::is_html_output()) { "Figure \\@ref(fig:branch) c"} else { "Figure \\@ref(fig:branch-proj) c"})`). The underlying geometry lies in the first two dimensions, while the remaining $(p-2)$ dimensions contain Gaussian noise.

Branch $1$ is generated from $X_1 \sim U(0, 1)$, and $X_2 = 0.1X_1 + X_1^2 + \epsilon$, where $\epsilon \sim U(0, 0.05)$. This produces a gently upward-curving parabola in the right half-plane. Branch $2$ is generated from $X_{1} \sim U(-1, 0)$, and $X_2 = 0.1X_{1} - 2X_{1}^2 + \epsilon$, where $\epsilon \sim U(0, 0.05)$. This creates a steeper, leftward-facing curve in the left half-plane. Branches $3$ to $k$ are then added iteratively. Each new branch $i$ begins at a randomly chosen starting point $(x_{\text{start}}, y_{\text{start}})$ from the existing structure, where the horizontal coordinate $x_{\text{start}}$ is restricted to $[-0.15, 0.15]$ to ensure connectivity with the initial branches. From this starting point, the branch extends over a unit-length interval $X_1 \in [x_{\text{start}}, x_{\text{start}} + 1]$, with structure $X_2 = 0.1X_1 - s_i(X_1^2 - x_{\text{start}}) + y_{\text{start}}, ; s_i \in {-2, -1.5, -1, -0.5, 0, 0.5, 1.5}$. Here, each branch $i$ is assigned a single scale factor $s_i$, sampled independently from the set, which controls the curvature of that branch. The parameter $(x_{\text{start}}, y_{\text{start}})$ is the starting point chosen from the existing structure. This construction yields $k$ spatially connected, nonlinear branches with varying curvature. For $p > 2$, Gaussian noise $X_j \sim N(0, 0.05^2)$ is added to embed the $2\text{-}D$ branches into $p\text{-}D$, where $j = 3, \dots, p$.

#### `gen_orglinearbranches()`

The `gen_orglinearbranches(n, p, k)` function generates a dataset of $n$ points forming $k$ approximately linear branches embedded in $p\text{-}D$ (`r knitr::asis_output(if (knitr::is_html_output()) { "Figure \\@ref(fig:branch) d"} else { "Figure \\@ref(fig:branch-proj) d"})`). Each branch lies within a distinct $2\text{-}D$ subspace, while the remaining $p - 2$ dimensions contain Gaussian noise.

To construct each branch, a unique or repeated pair of dimensions is selected from the $\binom{p}{2}$ possible combinations of dimensions. If $k \leq \binom{p}{2}$, combinations are sampled without replacement. If $k > \binom{p}{2}$, additional pairs are sampled with replacement to reach $k$ total branches. Each selected pair $(i_1, i_2)$ defines the $2\text{-}D$ plane for branch $i$.

For branch $i$, $n_i$ points are generated, where $\sum_{i=1}^k n_i = n$. The structure follows: $X_{i_1} \sim U(0, 2)$, and $X_{i_2} = -s_iX_{i_1} + \epsilon, \quad \epsilon \sim U(0, 0.5)$, where $s_i$ is a scale factor controlling the slope. When $k \leq \binom{p}{2}$, $s_i = 1$. When sampling with replacement, $s_i$ is drawn from the set ${1, 1.5, 2, \dots, 8}$ in increments of $0.5$. For $p > 2$, the remaining dimensions contain independent Gaussian noise: $X_j \sim N(0, 0.1^2)$, for $j \notin \{i_1, i_2\}$.

@@

#### `gen_orgcurvybranches()`

The `gen_orgcurvybranches(n, p, k)` function generates a dataset of $n$ points forming $k$ curvilinear branches embedded in a $p$-dimensional space (`r knitr::asis_output(if (knitr::is_html_output()) { "Figure \\@ref(fig:branch) e"} else { "Figure \\@ref(fig:branch-proj) e"})`). Each branch is constructed in a unique or repeated $2\text{-}D$ subspace of the $p$-dimensional space, with curvature induced by a second-degree polynomial structure. The remaining $p - 2$ dimensions contain Gaussian noise.

Let $\binom{p}{2}$ denote the number of unique $2\text{-}D$ subspace combinations. When $k \leq \binom{p}{2}$, $k$ distinct subspace pairs $(i_1, i_2)$ are sampled without replacement. Otherwise, combinations are selected with replacement to reach $k$ total branches. For each branch $i = 1, \dots, k$, a random scale factor $s_i$ is used to vary the curvature; If $k \leq \binom{p}{2}$: $s_i = 1$, or If $k > \binom{p}{2}$: $s_i \in \{1, 1.5, \dots, 8\}$.

Each branch contains $n_i$ points such that $\sum_{i=1}^k n_i = n$, where the vector $(n_1, \dots, n_k)$ is randomly drawn using the helper function `gen_nsum()` to partition $n$. Within its assigned subspace, branch $i$ is defined by: $X_{i_1} \sim U(0, 2)$, and $X_{i_2} = -s_iX_{i_1}^2 + \epsilon, \quad \epsilon \sim U(0, 0.5)$. This forms a smooth downward-opening parabola in the plane defined by $(X_{i_1}, X_{i_2})$, with the degree of curvature controlled by $s_i$. For $p > 2$, all remaining dimensions are independent Gaussian noise: $X_j \sim N(0, 0.1^2), \quad \text{for } j \notin \{i_1, i_2\}$.

@@

>>>>

#### Branching

@@

The simplest structures are approximately linear branches, generated by the `gen_linearbranches(n, p, k)` function (`r knitr::asis_output(if (knitr::is_html_output()) { "Figure \\@ref(fig:branch) b"} else { "Figure \\@ref(fig:branch-proj) b"})`). These consist of $k$ short line segments in the first two dimensions, with added jitter to simulate variability. Mathematically, each branch $i$ is defined as

$$
X_1 \sim U(a_i, b_i), \quad X_2 = s_i (X_1 - x_{\text{start},i}) + y_{\text{start},i} + \epsilon, \quad \epsilon \sim U(0, \delta),
$$

where $(x_{\text{start},i}, y_{\text{start},i})$ is the starting point of branch $i$, $\delta$ controls local jitter, and $s_i$ is the slope, initialized as

$$
s_i =
\begin{cases}
0.5 & i = 1, \\
-0.5 & i = 2, \\
\text{randomly sampled from } [s_{\min}, s_{\max}] & i = 3, \dots, k.
\end{cases}
$$

Branches $1$ and $2$ are initialized with fixed slopes and intercepts, while later branches are iteratively added at locations chosen to avoid overlap with existing branches, producing a set of connected linear paths.

To introduce curvature, the `gen_curvybranches(n, p, k)` function generates $k$ curvilinear branches in $p\text{-}D$ (`r knitr::asis_output(if (knitr::is_html_output()) { "Figure \\@ref(fig:branch) c"} else { "Figure \\@ref(fig:branch-proj) c"})`). Branches $1$ and $2$ are simple parabolas defined as

$$
\begin{aligned}
\text{Branch 1: } & X_1 \sim U(0,1), \quad X_2 = 0.1 X_1 + X_1^2 + \epsilon, \\
\text{Branch 2: } & X_1 \sim U(-1,0), \quad X_2 = 0.1 X_1 - 2 X_1^2 + \epsilon, \quad \epsilon \sim U(0, \delta),
\end{aligned}
$$

where $\delta$ controls local jitter. Additional branches are attached iteratively to existing structures. Each new branch $i$ starts at a selected point $(x_{\text{start},i}, y_{\text{start},i})$ from the current structure and extends according to

$$
X_1 \sim U(x_{\text{start},i}, x_{\text{start},i}+1), \quad X_2 = 0.1 X_1 - s_i (X_1^2 - x_{\text{start},i}) + y_{\text{start},i},
$$

where $s_i$ is a scale factor controlling the curvature of branch $i$. For the first few initial branches, $s_i$ can be fixed (e.g., $s_1 = 1, s_2 = 2$), while for subsequent branches it is sampled from a predefined set, such as $s_i \in \{-2, -1.5, -1, -0.5, 0, 0.5, 1, 1.5\}$, to create variability in curvature.

The `gen_expbranches(n, p, k)` function creates $k$ exponential branches in $2\text{-}D$, radiating from a central region (`r knitr::asis_output(if (knitr::is_html_output()) { "Figure \\@ref(fig:branch) a"} else { "Figure \\@ref(fig:branch-proj) a"})`). Each branch $i$ is defined as

$$
X_1 \sim U(-2,2), \quad X_2 = \exp(\sigma_i \, s_i \, X_1) + \epsilon, \quad \epsilon \sim U(0, \delta), \quad s_i \sim U(0.5,2),
$$

where $\sigma_i = (-1)^{i+1}$ alternates the sign of the exponent to produce mirror-symmetric branches. The parameter $s_i$ controls the steepness of branch $i$, and $\delta$ introduces small local jitter.

@@

High-dimensional generalizations are provided by `gen_orglinearbranches(n, p, k)` and `gen_orgcurvybranches(n, p, k)` (`r knitr::asis_output(if (knitr::is_html_output()) { "Figure \\@ref(fig:branch) d-e"} else { "Figure \\@ref(fig:branch-proj) d-e"})`). Each branch is embedded in a unique or repeated $2\text{-}D$ subspace of the $p\text{-}D$ space. Linear branches follow

$$
X_{i_1} \sim U(a_i,b_i), \quad X_{i_2} = s_i X_{i_1} + \epsilon, \quad \epsilon \sim N(0, \sigma^2),
$$

while curvilinear branches include a quadratic term

$$
X_{i_1} \sim U(a_i,b_i), \quad X_{i_2} = - s_i X_{i_1}^2 + \epsilon, \quad \epsilon \sim N(0, \sigma^2),
$$

where $a_i, b_i$ define the range of the first coordinate for branch $i$, and $\epsilon$ is Gaussian noise added to introduce variability. The scale factor $s_i$ controls slope (linear branches) or curvature (curvilinear branches) and is assigned as follows: for the first $\binom{p}{2}$ branches, $s_i = 1$; for additional branches when $k > \binom{p}{2}$, $s_i$ is randomly drawn from the set $\{1, 1.5, 2, \dots, 8\}$.

@@

@@

For all branch-generating functions, when the target dimensionality $p > 2$, additional dimensions are filled with independent noise, allowing the $2\text{-}D$ or $2\text{-}D$-subspace structures to be naturally embedded in higher-dimensional space while preserving the intended geometric patterns.

### 2025-09-20 paper-cardinalR.tex
Updated branching section

\subsubsection{Branching}\label{branching}

@@

\caption{\label{tab:branching-tb-pdf}cardinalR branching data generation functions}
\centering
\begin{tabular}[t]{>{\raggedright\arraybackslash}p{4cm}>{\raggedright\arraybackslash}p{8cm}}
\toprule
Function & Explanation\\
\midrule
gen\_expbranches & Generate a structure with exponential shaped branches.\\
gen\_linearbranches & Generate a structure with linear shaped branches.\\
gen\_curvybranches & Generate a structure with curvy shaped branches.\\
gen\_orglinearbranches & Generate a structure with linear shaped branches originated in one point.\\
gen\_orgcurvybranches & Generate a structure with curvy shaped branches originated in one point.\\
\bottomrule
\end{tabular}
\end{table}

@@

\paragraph{\texorpdfstring{\texttt{gen\_expbranches()}}{gen\_expbranches()}}\label{gen_expbranches}

The \texttt{gen\_expbranches(n,\ p,\ k)} function generates a dataset of \(n\) points forming \(k\) exponential branches in \(2\text{-}D\), with optional noise dimensions to embed the structure in \(p\text{-}D\) (Figure \ref{fig:branch-proj} a). These branches grow in opposite directions, producing a radiating curvilinear structure from a central region.

Each branch \(i\) is constructed using \(X_1 \sim U(-2, 2)\) and \(X_2 = \exp(\pm s_iX_1) + \epsilon\), where: \(\epsilon \sim U(0, 0.1)\) adds local jitter, \(s_i \in [0.5, 2]\) is randomly sampled for each branch. The sign of the exponent alternates between branches: odd-numbered branches have a negative exponent (decaying pattern), and even-numbered branches have a positive exponent (growing pattern), producing mirror-symmetric curves with varying steepness. For \(p > 2\), Gaussian noise \(X_j \sim N(0, 0.1^2)\) is added to embed the \(2\text{-}D\) branches into \(p\text{-}D\), where \(j = 3, \dots, p\).

@@

\paragraph{\texorpdfstring{\texttt{gen\_linearbranches()}}{gen\_linearbranches()}}\label{gen_linearbranches}

The \texttt{gen\_linearbranches(n,\ p,\ k)} function generates a dataset of \(n\) points forming \(k\) approximately linear branches in \(p\text{-}D\) (Figure \ref{fig:branch-proj} b). The core structure lies in the first two dimensions and additional dimensions carry Gaussian noise.

The \texttt{gen\_linearbranches(n,\ p,\ k)} function generates a dataset of \(n\) points in \(p\text{-}D\) forming \(k\) linear branches. Each branch is a line segment with added jitter to simulate noise. Branches \(1\) and \(2\) are initialized with fixed slopes and intercepts: branch \(1\) has slope \(0.5\) and intercept \(0\), with \(X_1 \sim U(-2, 8)\) and \(X_2 = 0.5 X_1 + \epsilon\), \(\epsilon \sim U(0, 0.5)\); branch \(2\) has slope \(-0.5\) and intercept \(0\), with \(X_1 \sim U(-6, 2)\) and \(X_2 = -0.5 X_1 + \epsilon\), \(\epsilon \sim U(0, 0.5)\).

Branches \(3\) to \(k\) are added iteratively. Each new branch starts at a location outside predefined exclusion zones to avoid overlap with the first two branches. Specifically, \(X_1\) starting points are sampled outside the intervals \([-8, -7]\), \([-2, 2]\), and \([7, 8]\), and \(X_2\) starting points are sampled outside \([7, 8]\). For each new branch \(i\), \(X_1\) values are drawn over a short interval \([x_{\text{start}}, x_{\text{start}} + 1]\), and \(X_2\) values are calculated as \(X_2 = s_i (X_1 - x_{\text{start}}) + y_{\text{start}} + \epsilon\), where \(s_i\) is a randomly chosen slope from a filtered sequence and \(\epsilon \sim U(0, 0.2)\). For \(p > 2\), Gaussian noise \(X_j \sim N(0, 0.05^2)\) is added to embed the \(2\text{-}D\) branches into \(p\text{-}D\), for \(j = 3, \dots, p\).

\paragraph{\texorpdfstring{\texttt{gen\_curvybranches()}}{gen\_curvybranches()}}\label{gen_curvybranches}

The \texttt{gen\_curvybranches(n,\ p,\ k)} function generates a dataset of \(n\) points forming \(k\) curvilinear branches embedded in \(p\text{-}D\) (Figure \ref{fig:branch-proj} c). The underlying geometry lies in the first two dimensions, while the remaining \((p-2)\) dimensions contain Gaussian noise.

Branch \(1\) is generated from \(X_1 \sim U(0, 1)\), and \(X_2 = 0.1X_1 + X_1^2 + \epsilon\), where \(\epsilon \sim U(0, 0.05)\). This produces a gently upward-curving parabola in the right half-plane. Branch \(2\) is generated from \(X_{1} \sim U(-1, 0)\), and \(X_2 = 0.1X_{1} - 2X_{1}^2 + \epsilon\), where \(\epsilon \sim U(0, 0.05)\). This creates a steeper, leftward-facing curve in the left half-plane. Branches \(3\) to \(k\) are then added iteratively. Each new branch \(i\) begins at a randomly chosen starting point \((x_{\text{start}}, y_{\text{start}})\) from the existing structure, where the horizontal coordinate \(x_{\text{start}}\) is restricted to \([-0.15, 0.15]\) to ensure connectivity with the initial branches. From this starting point, the branch extends over a unit-length interval \(X_1 \in [x_{\text{start}}, x_{\text{start}} + 1]\), with structure \(X_2 = 0.1X_1 - s_i(X_1^2 - x_{\text{start}}) + y_{\text{start}}, ; s_i \in {-2, -1.5, -1, -0.5, 0, 0.5, 1.5}\). Here, each branch \(i\) is assigned a single scale factor \(s_i\), sampled independently from the set, which controls the curvature of that branch. The parameter \((x_{\text{start}}, y_{\text{start}})\) is the starting point chosen from the existing structure. This construction yields \(k\) spatially connected, nonlinear branches with varying curvature. For \(p > 2\), Gaussian noise \(X_j \sim N(0, 0.05^2)\) is added to embed the \(2\text{-}D\) branches into \(p\text{-}D\), where \(j = 3, \dots, p\).

\paragraph{\texorpdfstring{\texttt{gen\_orglinearbranches()}}{gen\_orglinearbranches()}}\label{gen_orglinearbranches}

The \texttt{gen\_orglinearbranches(n,\ p,\ k)} function generates a dataset of \(n\) points forming \(k\) approximately linear branches embedded in \(p\text{-}D\) (Figure \ref{fig:branch-proj} d). Each branch lies within a distinct \(2\text{-}D\) subspace, while the remaining \(p - 2\) dimensions contain Gaussian noise.

To construct each branch, a unique or repeated pair of dimensions is selected from the \(\binom{p}{2}\) possible combinations of dimensions. If \(k \leq \binom{p}{2}\), combinations are sampled without replacement. If \(k > \binom{p}{2}\), additional pairs are sampled with replacement to reach \(k\) total branches. Each selected pair \((i_1, i_2)\) defines the \(2\text{-}D\) plane for branch \(i\).

For branch \(i\), \(n_i\) points are generated, where \(\sum_{i=1}^k n_i = n\). The structure follows: \(X_{i_1} \sim U(0, 2)\), and \(X_{i_2} = -s_iX_{i_1} + \epsilon, \quad \epsilon \sim U(0, 0.5)\), where \(s_i\) is a scale factor controlling the slope. When \(k \leq \binom{p}{2}\), \(s_i = 1\). When sampling with replacement, \(s_i\) is drawn from the set \({1, 1.5, 2, \dots, 8}\) in increments of \(0.5\). For \(p > 2\), the remaining dimensions contain independent Gaussian noise: \(X_j \sim N(0, 0.1^2)\), for \(j \notin \{i_1, i_2\}\).

@@

\paragraph{\texorpdfstring{\texttt{gen\_orgcurvybranches()}}{gen\_orgcurvybranches()}}\label{gen_orgcurvybranches}

The \texttt{gen\_orgcurvybranches(n,\ p,\ k)} function generates a dataset of \(n\) points forming \(k\) curvilinear branches embedded in a \(p\)-dimensional space (Figure \ref{fig:branch-proj} e). Each branch is constructed in a unique or repeated \(2\text{-}D\) subspace of the \(p\)-dimensional space, with curvature induced by a second-degree polynomial structure. The remaining \(p - 2\) dimensions contain Gaussian noise.

Let \(\binom{p}{2}\) denote the number of unique \(2\text{-}D\) subspace combinations. When \(k \leq \binom{p}{2}\), \(k\) distinct subspace pairs \((i_1, i_2)\) are sampled without replacement. Otherwise, combinations are selected with replacement to reach \(k\) total branches. For each branch \(i = 1, \dots, k\), a random scale factor \(s_i\) is used to vary the curvature; If \(k \leq \binom{p}{2}\): \(s_i = 1\), or If \(k > \binom{p}{2}\): \(s_i \in \{1, 1.5, \dots, 8\}\).

Each branch contains \(n_i\) points such that \(\sum_{i=1}^k n_i = n\), where the vector \((n_1, \dots, n_k)\) is randomly drawn using the helper function \texttt{gen\_nsum()} to partition \(n\). Within its assigned subspace, branch \(i\) is defined by: \(X_{i_1} \sim U(0, 2)\), and \(X_{i_2} = -s_iX_{i_1}^2 + \epsilon, \quad \epsilon \sim U(0, 0.5)\). This forms a smooth downward-opening parabola in the plane defined by \((X_{i_1}, X_{i_2})\), with the degree of curvature controlled by \(s_i\). For \(p > 2\), all remaining dimensions are independent Gaussian noise: \(X_j \sim N(0, 0.1^2), \quad \text{for } j \notin \{i_1, i_2\}\).

@@

>>>>

\paragraph{Branching}\label{branching}

@@

\caption{\label{tab:branching-tb-pdf}cardinalR branching data generation functions}
\centering
\begin{tabular}[t]{>{\raggedright\arraybackslash}p{4cm}>{\raggedright\arraybackslash}p{8cm}}
\toprule
Function & Explanation\\
\midrule
gen\_expbranches & Exponential shaped branches.\\
gen\_linearbranches & Linear shaped branches.\\
gen\_curvybranches & Curvy shaped branches.\\
gen\_orglinearbranches & Linear shaped branches originated in one point.\\
gen\_orgcurvybranches & Curvy shaped branches originated in one point.\\
\bottomrule
\end{tabular}
\end{table}

@@

The simplest structures are approximately linear branches, generated by the \texttt{gen\_linearbranches(n,\ p,\ k)} function (Figure \ref{fig:branch-proj} b). These consist of \(k\) short line segments in the first two dimensions, with added jitter to simulate variability. Mathematically, each branch \(i\) is defined as

\[
X_1 \sim U(a_i, b_i), \quad X_2 = s_i (X_1 - x_{\text{start},i}) + y_{\text{start},i} + \epsilon, \quad \epsilon \sim U(0, \delta),
\]

where \((x_{\text{start},i}, y_{\text{start},i})\) is the starting point of branch \(i\), \(\delta\) controls local jitter, and \(s_i\) is the slope, initialized as

\[
s_i =
\begin{cases}
0.5 & i = 1, \\
-0.5 & i = 2, \\
\text{randomly sampled from } [s_{\min}, s_{\max}] & i = 3, \dots, k.
\end{cases}
\]

Branches \(1\) and \(2\) are initialized with fixed slopes and intercepts, while later branches are iteratively added at locations chosen to avoid overlap with existing branches, producing a set of connected linear paths.

To introduce curvature, the \texttt{gen\_curvybranches(n,\ p,\ k)} function generates \(k\) curvilinear branches in \(p\text{-}D\) (Figure \ref{fig:branch-proj} c). Branches \(1\) and \(2\) are simple parabolas defined as

\[
\begin{aligned}
\text{Branch 1: } & X_1 \sim U(0,1), \quad X_2 = 0.1 X_1 + X_1^2 + \epsilon, \\
\text{Branch 2: } & X_1 \sim U(-1,0), \quad X_2 = 0.1 X_1 - 2 X_1^2 + \epsilon, \quad \epsilon \sim U(0, \delta),
\end{aligned}
\]

where \(\delta\) controls local jitter. Additional branches are attached iteratively to existing structures. Each new branch \(i\) starts at a selected point \((x_{\text{start},i}, y_{\text{start},i})\) from the current structure and extends according to

\[
X_1 \sim U(x_{\text{start},i}, x_{\text{start},i}+1), \quad X_2 = 0.1 X_1 - s_i (X_1^2 - x_{\text{start},i}) + y_{\text{start},i},
\]

where \(s_i\) is a scale factor controlling the curvature of branch \(i\). For the first few initial branches, \(s_i\) can be fixed (e.g., \(s_1 = 1, s_2 = 2\)), while for subsequent branches it is sampled from a predefined set, such as \(s_i \in \{-2, -1.5, -1, -0.5, 0, 0.5, 1, 1.5\}\), to create variability in curvature.

The \texttt{gen\_expbranches(n,\ p,\ k)} function creates \(k\) exponential branches in \(2\text{-}D\), radiating from a central region (Figure \ref{fig:branch-proj} a). Each branch \(i\) is defined as

\[
X_1 \sim U(-2,2), \quad X_2 = \exp(\sigma_i \, s_i \, X_1) + \epsilon, \quad \epsilon \sim U(0, \delta), \quad s_i \sim U(0.5,2),
\]

where \(\sigma_i = (-1)^{i+1}\) alternates the sign of the exponent to produce mirror-symmetric branches. The parameter \(s_i\) controls the steepness of branch \(i\), and \(\delta\) introduces small local jitter.

@@

High-dimensional generalizations are provided by \texttt{gen\_orglinearbranches(n,\ p,\ k)} and \texttt{gen\_orgcurvybranches(n,\ p,\ k)} (Figure \ref{fig:branch-proj} d-e). Each branch is embedded in a unique or repeated \(2\text{-}D\) subspace of the \(p\text{-}D\) space. Linear branches follow

\[
X_{i_1} \sim U(a_i,b_i), \quad X_{i_2} = s_i X_{i_1} + \epsilon, \quad \epsilon \sim N(0, \sigma^2),
\]

while curvilinear branches include a quadratic term

\[
X_{i_1} \sim U(a_i,b_i), \quad X_{i_2} = - s_i X_{i_1}^2 + \epsilon, \quad \epsilon \sim N(0, \sigma^2),
\]

where \(a_i, b_i\) define the range of the first coordinate for branch \(i\), and \(\epsilon\) is Gaussian noise added to introduce variability. The scale factor \(s_i\) controls slope (linear branches) or curvature (curvilinear branches) and is assigned as follows: for the first \(\binom{p}{2}\) branches, \(s_i = 1\); for additional branches when \(k > \binom{p}{2}\), \(s_i\) is randomly drawn from the set \(\{1, 1.5, 2, \dots, 8\}\).

@@

@@

For all branch-generating functions, when the target dimensionality \(p > 2\), additional dimensions are filled with independent noise, allowing the \(2\text{-}D\) or \(2\text{-}D\)-subspace structures to be naturally embedded in higher-dimensional space while preserving the intended geometric patterns.
